Reanalysis Step 2: Multiple Imputation (Clean Version)

### 2025-08-30 reanalysis-2-multiple_imputation.qmd
predictor matrix fixes

```{r setup-predictors}
# Use quickpred to get smart predictor selection
# This selects predictors with correlation >= 0.10 (R^2 >= 0.01)
# As recommended in Van Buuren's MICE documentation

# Variables that should always be included as predictors (key variables)
vars_always_predict <- c(
  "ever_lapse_binary",      # Exposure
  "baseline_VA_logMAR",     # Key outcome
  "outcome_VA_logMAR",      # Key outcome
  "age",                    # Demographics
  "gender_cat",
  "race_ethnic_cat",        # Demographics
  "insurance_cat",          # Demographics/SES
  "CCI", "DCSI",           # Comorbidities
  "person_dr",              # DR severity - key clinical variable
  "anti_VEGF",              # Key treatment
  "focal_laser_flag",       # Treatment
  "PRP_flag"                # Treatment
)

# Variables that should never be predictors
vars_never_predict <- c(
  vars_no_impute_no_predict
)

# Generate predictor matrix using quickpred
# Lower threshold to 0.05 to capture more predictive relationships
# This includes variables with correlations >= 0.05 (R^2 >= 0.0025)
pred_matrix <- quickpred(
  df_mice, 
  exclude = vars_never_predict,
  include = vars_always_predict,
  mincor = 0.05,  # Lowered from 0.10 to include more predictors
  method = "pearson"
)

# No multilevel imputation: ensure cluster ID is not used as predictor
pred_matrix[, cluster_var] <- 0
pred_matrix[cluster_var, ] <- 0

# Ensure temporal ordering for VA variables
# Baseline can predict intermediate and outcome
pred_matrix["intermediate_VA_logMAR", "baseline_VA_logMAR"] <- 1
pred_matrix["outcome_VA_logMAR", "baseline_VA_logMAR"] <- 1

# Intermediate can predict outcome  
pred_matrix["outcome_VA_logMAR", "intermediate_VA_logMAR"] <- 1

# Prevent reverse-time prediction
pred_matrix["baseline_VA_logMAR", "outcome_VA_logMAR"] <- 0
pred_matrix["baseline_VA_logMAR", "intermediate_VA_logMAR"] <- 0
pred_matrix["intermediate_VA_logMAR", "outcome_VA_logMAR"] <- 0

# Add theoretically important relationships even if correlation is low
# Demographics should predict each other
demographic_vars <- c("age", "gender_cat", "race_ethnic_cat", "insurance_cat")
for (v1 in demographic_vars) {
  for (v2 in demographic_vars) {
    if (v1 != v2 && v1 %in% rownames(pred_matrix) && v2 %in% colnames(pred_matrix)) {
      pred_matrix[v1, v2] <- 1
    }
  }
}

# Comorbidities should predict DR severity and vice versa
if ("person_dr" %in% rownames(pred_matrix)) {
  if ("CCI" %in% colnames(pred_matrix)) pred_matrix["person_dr", "CCI"] <- 1
  if ("DCSI" %in% colnames(pred_matrix)) pred_matrix["person_dr", "DCSI"] <- 1
  if ("glaucoma_bef_hitplus_cat" %in% colnames(pred_matrix)) pred_matrix["person_dr", "glaucoma_bef_hitplus_cat"] <- 1
}

# Treatment variables should predict each other (treatment patterns)
treatment_vars <- c("anti_VEGF", "focal_laser_flag", "PRP_flag", "other_inject")
for (v1 in treatment_vars) {
  for (v2 in treatment_vars) {
    if (v1 != v2 && v1 %in% rownames(pred_matrix) && v2 %in% colnames(pred_matrix)) {
      pred_matrix[v1, v2] <- 1
    }
  }
}

# Ensure demographics predict VA timepoints and DR severity
va_vars <- c("baseline_VA_logMAR", "intermediate_VA_logMAR", "outcome_VA_logMAR", "person_dr")
for (va in va_vars) {
  for (demo in demographic_vars) {
    if (va %in% rownames(pred_matrix) && demo %in% colnames(pred_matrix)) {
      pred_matrix[va, demo] <- 1
    }
  }
}

# Ensure ocular comorbidities and cataract surgery predict VA and DR severity
ocular_vars <- c("glaucoma_bef_hitplus_cat", "otherretina_bef_hitplus_cat", "catsurg_before_hitplus_cat")
for (va in va_vars) {
  for (ov in ocular_vars) {
    if (va %in% rownames(pred_matrix) && ov %in% colnames(pred_matrix)) {
      pred_matrix[va, ov] <- 1
    }
  }
}

# Ensure treatments predict intermediate and outcome VA
for (va in c("intermediate_VA_logMAR", "outcome_VA_logMAR")) {
  for (tr in treatment_vars) {
    if (va %in% rownames(pred_matrix) && tr %in% colnames(pred_matrix)) {
      pred_matrix[va, tr] <- 1
    }
  }
}

# Fellow-eye predictors with weight=2 for stronger influence
fellow_vars <- c("fellow_baseline_VA", "fellow_intermediate_VA", "fellow_outcome_VA", "fellow_dr")
for (v in fellow_vars) {
  if (v %in% rownames(pred_matrix)) {
    pred_matrix[v, ] <- 0
  }
}
if ("fellow_baseline_VA" %in% colnames(pred_matrix)) {
  pred_matrix["baseline_VA_logMAR", "fellow_baseline_VA"] <- 1
}
if ("fellow_intermediate_VA" %in% colnames(pred_matrix)) {
  pred_matrix["intermediate_VA_logMAR", "fellow_intermediate_VA"] <- 1
}
if ("fellow_outcome_VA" %in% colnames(pred_matrix)) {
  pred_matrix["outcome_VA_logMAR", "fellow_outcome_VA"] <- 1
}
if ("fellow_dr" %in% colnames(pred_matrix) && "person_dr" %in% rownames(pred_matrix)) {
  pred_matrix["person_dr", "fellow_dr"] <- 1
}

log_message("Predictor matrix configured with temporal constraints")

# Print summary of predictor counts for each variable
pred_counts <- data.frame(
  Variable = rownames(pred_matrix),
  N_Predictors = rowSums(pred_matrix > 0),
  N_Predicting = colSums(pred_matrix > 0)[rownames(pred_matrix)]
)
pred_counts <- pred_counts[order(-pred_counts$N_Predictors), ]

log_message("\nTop variables by number of predictors:")
print(head(pred_counts, 15))
```

>>>>

```{r setup-predictors}
# Create predictor matrix based on theoretical relationships with strict temporal ordering
# Start with empty matrix (no predictors)
pred_matrix <- matrix(0, nrow = ncol(df_mice), ncol = ncol(df_mice))
colnames(pred_matrix) <- names(df_mice)
rownames(pred_matrix) <- names(df_mice)

# STEP 1: Ensure complete exclusions for non-imputed variables
# e_mrn_deidentified and row_id should be completely excluded (rows and columns = 0)
if ("e_mrn_deidentified" %in% names(df_mice)) {
  pred_matrix[, "e_mrn_deidentified"] <- 0
  pred_matrix["e_mrn_deidentified", ] <- 0
}
if ("row_id" %in% names(df_mice)) {
  pred_matrix[, "row_id"] <- 0
  pred_matrix["row_id", ] <- 0
}

# Ensure ever_lapse_binary and cohort_id are never imputed (rows = 0) but can be predictors
if ("ever_lapse_binary" %in% rownames(pred_matrix)) {
  pred_matrix["ever_lapse_binary", ] <- 0  # Never imputed
}
if ("cohort_id" %in% rownames(pred_matrix)) {
  pred_matrix["cohort_id", ] <- 0  # Never imputed
}

# Ensure all fellow variables are never imputed (rows = 0)
fellow_vars <- c("fellow_baseline_VA", "fellow_intermediate_VA", "fellow_outcome_VA", "fellow_dr", "n_eyes")
for (v in fellow_vars) {
  if (v %in% rownames(pred_matrix)) {
    pred_matrix[v, ] <- 0  # Never imputed
  }
}

# Define variable groups for clarity
demographic_vars <- c("age", "gender_cat", "race_ethnic_cat", "insurance_cat")
global_comorbidity_vars <- c("CCI", "DCSI")
ocular_comorbidity_vars <- c("glaucoma_bef_hitplus_cat", "otherretina_bef_hitplus_cat", "catsurg_before_hitplus_cat")
treatment_vars <- c("anti_VEGF", "focal_laser_flag", "PRP_flag", "other_inject")

# STEP 2: Configure baseline_VA_logMAR predictors
# Predictors: demographics, global comorbidities, ocular comorbidities, exposure, fellow_baseline_VA
# NO treatments, NO intermediate/outcome VAs
if ("baseline_VA_logMAR" %in% rownames(pred_matrix)) {
  # Demographics
  for (v in demographic_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["baseline_VA_logMAR", v] <- 1
  }
  # Global comorbidities
  for (v in global_comorbidity_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["baseline_VA_logMAR", v] <- 1
  }
  # Ocular comorbidities and cataract
  for (v in ocular_comorbidity_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["baseline_VA_logMAR", v] <- 1
  }
  # Exposure
  if ("ever_lapse_binary" %in% colnames(pred_matrix)) pred_matrix["baseline_VA_logMAR", "ever_lapse_binary"] <- 1
  # Fellow eye baseline only
  if ("fellow_baseline_VA" %in% colnames(pred_matrix)) pred_matrix["baseline_VA_logMAR", "fellow_baseline_VA"] <- 1

  # Explicitly exclude treatments and future timepoints
  for (v in treatment_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["baseline_VA_logMAR", v] <- 0
  }
  if ("intermediate_VA_logMAR" %in% colnames(pred_matrix)) pred_matrix["baseline_VA_logMAR", "intermediate_VA_logMAR"] <- 0
  if ("outcome_VA_logMAR" %in% colnames(pred_matrix)) pred_matrix["baseline_VA_logMAR", "outcome_VA_logMAR"] <- 0
  if ("fellow_intermediate_VA" %in% colnames(pred_matrix)) pred_matrix["baseline_VA_logMAR", "fellow_intermediate_VA"] <- 0
  if ("fellow_outcome_VA" %in% colnames(pred_matrix)) pred_matrix["baseline_VA_logMAR", "fellow_outcome_VA"] <- 0
}

# STEP 3: Configure intermediate_VA_logMAR predictors
# Predictors: baseline_VA, demographics, CCI/DCSI, ocular comorbidities, exposure, fellow_baseline_VA
# NO treatments, NO fellow_intermediate_VA, NO outcome VAs
if ("intermediate_VA_logMAR" %in% rownames(pred_matrix)) {
  # Baseline VA
  if ("baseline_VA_logMAR" %in% colnames(pred_matrix)) pred_matrix["intermediate_VA_logMAR", "baseline_VA_logMAR"] <- 1
  # Demographics
  for (v in demographic_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["intermediate_VA_logMAR", v] <- 1
  }
  # Global comorbidities
  for (v in global_comorbidity_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["intermediate_VA_logMAR", v] <- 1
  }
  # Ocular comorbidities and cataract
  for (v in ocular_comorbidity_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["intermediate_VA_logMAR", v] <- 1
  }
  # Exposure
  if ("ever_lapse_binary" %in% colnames(pred_matrix)) pred_matrix["intermediate_VA_logMAR", "ever_lapse_binary"] <- 1
  # Fellow eye baseline only (NOT intermediate)
  if ("fellow_baseline_VA" %in% colnames(pred_matrix)) pred_matrix["intermediate_VA_logMAR", "fellow_baseline_VA"] <- 1

  # Explicitly exclude treatments, fellow_intermediate_VA, and outcome VAs
  for (v in treatment_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["intermediate_VA_logMAR", v] <- 0
  }
  if ("fellow_intermediate_VA" %in% colnames(pred_matrix)) pred_matrix["intermediate_VA_logMAR", "fellow_intermediate_VA"] <- 0
  if ("outcome_VA_logMAR" %in% colnames(pred_matrix)) pred_matrix["intermediate_VA_logMAR", "outcome_VA_logMAR"] <- 0
  if ("fellow_outcome_VA" %in% colnames(pred_matrix)) pred_matrix["intermediate_VA_logMAR", "fellow_outcome_VA"] <- 0
}

# STEP 4: Configure outcome_VA_logMAR predictors
# Predictors: baseline_VA, intermediate_VA, demographics, CCI/DCSI, ocular comorbidities, exposure, treatments
# Fellow predictors: fellow_baseline_VA and fellow_outcome_VA (NOT fellow_intermediate_VA)
if ("outcome_VA_logMAR" %in% rownames(pred_matrix)) {
  # Prior timepoints
  if ("baseline_VA_logMAR" %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", "baseline_VA_logMAR"] <- 1
  if ("intermediate_VA_logMAR" %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", "intermediate_VA_logMAR"] <- 1
  # Demographics
  for (v in demographic_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", v] <- 1
  }
  # Global comorbidities
  for (v in global_comorbidity_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", v] <- 1
  }
  # Ocular comorbidities and cataract
  for (v in ocular_comorbidity_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", v] <- 1
  }
  # Exposure
  if ("ever_lapse_binary" %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", "ever_lapse_binary"] <- 1
  # Treatments (allowed for outcome)
  for (v in treatment_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", v] <- 1
  }
  # Fellow predictors: baseline and outcome only
  if ("fellow_baseline_VA" %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", "fellow_baseline_VA"] <- 1
  if ("fellow_outcome_VA" %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", "fellow_outcome_VA"] <- 1

  # Explicitly exclude fellow_intermediate_VA
  if ("fellow_intermediate_VA" %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", "fellow_intermediate_VA"] <- 0
}

# STEP 5: Configure person_dr predictors
# Predictors: demographics, CCI/DCSI, ocular comorbidities, baseline_VA, exposure, treatments, fellow_baseline_VA
# NO fellow_intermediate_VA, NO outcome_VA (to remain strictly pre-outcome)
if ("person_dr" %in% rownames(pred_matrix)) {
  # Demographics
  for (v in demographic_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["person_dr", v] <- 1
  }
  # Global comorbidities
  for (v in global_comorbidity_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["person_dr", v] <- 1
  }
  # Ocular comorbidities and cataract
  for (v in ocular_comorbidity_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["person_dr", v] <- 1
  }
  # Baseline VA
  if ("baseline_VA_logMAR" %in% colnames(pred_matrix)) pred_matrix["person_dr", "baseline_VA_logMAR"] <- 1
  # Exposure
  if ("ever_lapse_binary" %in% colnames(pred_matrix)) pred_matrix["person_dr", "ever_lapse_binary"] <- 1
  # Treatments
  for (v in treatment_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["person_dr", v] <- 1
  }
  # Fellow predictors: baseline only and fellow_dr
  if ("fellow_baseline_VA" %in% colnames(pred_matrix)) pred_matrix["person_dr", "fellow_baseline_VA"] <- 1
  if ("fellow_dr" %in% colnames(pred_matrix)) pred_matrix["person_dr", "fellow_dr"] <- 1

  # Explicitly exclude fellow_intermediate_VA and outcome_VA
  if ("fellow_intermediate_VA" %in% colnames(pred_matrix)) pred_matrix["person_dr", "fellow_intermediate_VA"] <- 0
  if ("outcome_VA_logMAR" %in% colnames(pred_matrix)) pred_matrix["person_dr", "outcome_VA_logMAR"] <- 0
  if ("fellow_outcome_VA" %in% colnames(pred_matrix)) pred_matrix["person_dr", "fellow_outcome_VA"] <- 0
}

# STEP 6: Configure other variables (demographics, comorbidities, treatments)
# Allow demographics to predict each other
for (v1 in demographic_vars) {
  for (v2 in demographic_vars) {
    if (v1 != v2 && v1 %in% rownames(pred_matrix) && v2 %in% colnames(pred_matrix)) {
      pred_matrix[v1, v2] <- 1
    }
  }
}

# CCI and DCSI predict each other and are predicted by demographics
if ("CCI" %in% rownames(pred_matrix)) {
  for (v in demographic_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["CCI", v] <- 1
  }
  if ("DCSI" %in% colnames(pred_matrix)) pred_matrix["CCI", "DCSI"] <- 1
  if ("ever_lapse_binary" %in% colnames(pred_matrix)) pred_matrix["CCI", "ever_lapse_binary"] <- 1
}
if ("DCSI" %in% rownames(pred_matrix)) {
  for (v in demographic_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["DCSI", v] <- 1
  }
  if ("CCI" %in% colnames(pred_matrix)) pred_matrix["DCSI", "CCI"] <- 1
  if ("ever_lapse_binary" %in% colnames(pred_matrix)) pred_matrix["DCSI", "ever_lapse_binary"] <- 1
}

# Cataract surgery predicted by demographics and comorbidities
if ("catsurg_before_hitplus_cat" %in% rownames(pred_matrix)) {
  for (v in demographic_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["catsurg_before_hitplus_cat", v] <- 1
  }
  for (v in global_comorbidity_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["catsurg_before_hitplus_cat", v] <- 1
  }
  if ("ever_lapse_binary" %in% colnames(pred_matrix)) pred_matrix["catsurg_before_hitplus_cat", "ever_lapse_binary"] <- 1
}

# Treatment variables predict each other (treatment patterns)
for (v1 in treatment_vars) {
  for (v2 in treatment_vars) {
    if (v1 != v2 && v1 %in% rownames(pred_matrix) && v2 %in% colnames(pred_matrix)) {
      pred_matrix[v1, v2] <- 1
    }
  }
}

log_message("Predictor matrix configured with strict temporal constraints")

# Verification: Print predictor counts for key variables
pred_counts <- data.frame(
  Variable = rownames(pred_matrix),
  N_Predictors = rowSums(pred_matrix > 0),
  N_Times_Used_As_Predictor = colSums(pred_matrix > 0)[rownames(pred_matrix)]
)
pred_counts <- pred_counts[order(-pred_counts$N_Predictors), ]

log_message("\nPredictor counts for all variables:")
print(pred_counts)

# Specific verification for variables with missingness
key_vars <- c("baseline_VA_logMAR", "intermediate_VA_logMAR", "outcome_VA_logMAR", "person_dr")
log_message("\nDetailed predictors for key variables with missingness:")
for (var in key_vars) {
  if (var %in% rownames(pred_matrix)) {
    predictors <- colnames(pred_matrix)[pred_matrix[var, ] == 1]
    log_message(paste0("\n", var, " (", length(predictors), " predictors):"))
    log_message(paste0("  ", paste(predictors, collapse = ", ")))
  }
}
```
